---
title: "Final Project"
subtitle: "INFO 526 Data Analysis and Visualization"
author: 
  - name: "Michael Burris"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "Final project for INFO 526 Data Analysis and Visualization"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false

# Load packages

#install.packages("pacman")

pacman::p_load(tidyverse
               )
```

## Data Set

```{r}
#| label: load-dataset
#| message: false

# Load dataset
er_data <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-08/care_state.csv")
glimpse(er_data)


```

*A brief description of your data set including its provenance, dimensions, etc. as well as the reason why you chose this data set.*

This data set comes from the Centers for Medicare and Medicaid Services (CMS) website. There are 1,232 rows and 8 columns of data. I chose this data set because I'm interested in working full-time in a Biotechnology or Healthcare-related industry, and this project presents a perfect opportunity for me to visualize and analyze health data.

*Make sure to load the data and use inline code for some of this information.*

## Questions

*The two questions you want to answer.*

1.  Which States are the most and least effective time wise at treating psychiatric/mental health patients that are admitted to the emergency room?
2.  At what point does the wait time lead to patients leaving at or above the national average?

## Analysis plan

*A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).*

- To answer the question about visit duration for psychiatric/mental health patients compared to non psychiatric/mental health patients, I plan to take the average visit time in minutes for both of these groups and compare the national averages as well as the averages by state. I will then visually display these minute values by state. The variables involved will be the condition, state, measure_id, and score. No external data will be needed to answer this question.
- To answer the question about wait times and at what point it leads to patients leaving above the national average, I will first calculate the average percentage of patients who come to the emergency room and leave before being seen. To best answer this question, I will need to create a new variable named "combined_score" that takes the average visit duration for psychiatric/mental health patients and the average visit duration of non-psychiatric/mental health patients, adds them together, and divides them by two. The reason I need this combined score is because I have no way of knowing which of these two categories the patients who left without being seen would fall into. I will also need the condition, state, measure_id, and score variables.

## Visuals

```{r}
# Which States are the most and least effective time wise at treating psychiatric/mental health patients that are admitted to the emergency room?

# Filter data to only include emergency room observations, patients that are non psyciatric/mental health patients, and only states not federal districts or territories.
patients <- er_data |>
  filter(condition == "Emergency Department", 
         measure_id == "OP_18c", state != "DC", state != "AS", state != "GU",
         state != "MP", state != "PR", state != "VI")

# Set my theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# Set code output width
options(width = 65)

# Set figure parameters
knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.68,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)

# Create data frame that only includes the 10 states with the longest ER visit times
longest_visits <- patients |>
  slice_max(score, n = 5)

# Create a data frame that only includes the 10 states with the shortest ER visit times
shortest_visits <- patients |>
  slice_min(score, n = 5)

# Join the two new data frames together
n = longest_visits |>
  full_join(shortest_visits)


# Plot my data
ggplot(n, aes(x = score, y = factor(state, levels = rev(levels(factor(state)))))) + 
  geom_point(size = 4) + 
  labs(
    title = "Highest & Lowest Median ER Visit Times",
    subtitle = "Psychiatric/mental health patients",
    x = "Minutes",
    y = NULL
  ) + 
  theme(
    plot.subtitle = element_text(color = "cornsilk4")
  )

# Sources
# https://dplyr.tidyverse.org/reference/filter.html
# https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States
# https://dplyr.tidyverse.org/reference/top_n.html
# https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/filter
# https://dplyr.tidyverse.org/reference/pull.html#ref-examples
# https://dplyr.tidyverse.org/reference/slice.html#ref-examples
# https://ggplot2.tidyverse.org/reference/geom_bar.html
# https://www.rdocumentation.org/packages/data.table/versions/1.18.4/topics/merge
# https://d2l.arizona.edu/d2l/le/content/1755726/viewContent/20559504/View
# https://r-graph-gallery.com/267-reorder-a-variable-in-ggplot2.html
# https://www.statology.org/ggplot-order-y-axis-alphabetically/
# https://medium.com/@MattOldach_65321/reversing-the-order-of-axis-in-a-ggplot2-scatterplot-b1d729b89118
# https://ggplot2.tidyverse.org/reference/geom_point.html#ref-examples


```

This visual highlights the 5 states that have the shortest and longest median ER visit times for psychiatric/mental health patients.
