---
title: "Final Project"
subtitle: "INFO 526 Data Analysis and Visualization"
author: 
  - name: "Michael Burris"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "Final project for INFO 526 Data Analysis and Visualization"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Data Set

```{r}
#| label: load-dataset
#| message: false

# Load packages
library(tidyverse)
library(here)

# Load dataset
er_data <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-08/care_state.csv")

# Add some more code here to show some important information about the data


```

*A brief description of your data set including its provenance, dimensions, etc. as well as the reason why you chose this data set.*

This data set comes from the Centers for Medicare and Medicaid Services (CMS) website. There are 1,232 rows and 8 columns of data. I chose this data set because I'm interested in working full-time in a Biotechnology or Healthcare-related role, and this project presents a perfect opportunity for me to visualize and analyze health data.

*Make sure to load the data and use inline code for some of this information.*

## Questions

The two questions you want to answer.

1.  How much time do psychiatric/mental health patients spend in the emergency room compared to non psychiatric/mental health patients?
2.  At what point does the wait time lead to patients leaving at or above the national average?

## Analysis plan

- *A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).*
- To answer the question about visit duration for psychiatric/mental health patients compared to non psychiatric/mental health patients, I plan to take the average visit time in minutes for both of these groups and compare the national averages as well as the averages by state. I will then visually display these minute values by state. The variables involved will be the condition, state, measure_id, and score. No external data will be needed to answer this question.
- To answer the question about wait times and at what point it leads to patients leaving above the national average, I will first calculate the average percentage of patients who come to the emergency room and leave before being seen. To best answer this question, I will need to create a new variable named "combined_score" that takes the average visit duration for psychiatric/mental health patients and the average visit duration of non-psychiatric/mental health patients, adds them together, and divides them by two. The reason I need this combined score is because I have no way of knowing which of these two categories the patients who left without being seen would fall into. I will also need the condition, state, measure_id, and score variables.

## Visuals

```{r}
# How efficient are emergency room departments at caring for psychiartic/mental healt patients?

# Visual
er_data |>
  filter(condition == "Emergency Department", measure_id == "OP_18b")

```
