---
title: "Emergency Room Care Across the United States"
subtitle: "INFO 526 Data Analysis and Visualization"
author: 
  - name: "Michael Burris"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "Final project for INFO 526 Data Analysis and Visualization"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

## Description

The data set used for this project comes from [TidyTuesday 2025 data sets.](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025#readme) Scroll down to Week 14 to view the Data and Source used to complete this project. Data comes from the Centers for Medicare and Medicaid Services (CMS) website. There are 1,232 rows and 8 columns of data.

```{=html}
<details>
<summary>Sources</summary>
# https://www.markdownguide.org/basic-syntax/
</details>
```

```{r}
#| label: load-pkgs
#| message: false

# Load packages
#install.packages("pacman")
pacman::p_load(tidyverse,
               scales,
               ggthemes
               )

# Theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# Code output width
options(width = 65)

# Figure parameters
knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.68,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)

```

## Data Set

```{r}
#| label: load-dataset
#| message: false

# Load dataset
er_data <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-08/care_state.csv")

# Show an overview of the data set
glimpse(er_data)

# Print off first 5 rows of data
head(er_data, n = 5)

```

```{=html}
<details>
<summary>Sources</summary>
# https://www.datacamp.com/doc/r/contents
</details>
```

I chose this data set because I'm interested in working full-time in a Biotechnology or Healthcare-related industry, and this project presents a perfect opportunity for me to visualize and analyze health data.

## Questions

1.  Which states/regions are the most and least effective time wise at treating psychiatric/mental health patients that are admitted to the emergency room?
2.  Does state population have an effect on what percent of patients leave the emergency room without being seen?

## Analysis plan

- To answer which states/regions are the most and least effective time wise at treating psychiatric/mental health patients admitted to the emergency room, I plan to create two visuals. I will need to use various variables from the care_state.csv file including condition, score, and measure_id. I will also need to utilize some external data that specifies the region for key states, states that have the shortest and longest median ER visit times for mental health patients. I will first visualize an overview of which states have the highest and lowest median visit times, then I will go a step further and visualize this by region as well.
- In order to address if state population has an effect on the percent of patients who leave the emergency room without being seen, I will need to use the condition, measure_id, and score variables. I will also need to pull in some external data to get population numbers for each state. I found an external source with population data estimates for each state for 2025, which I felt was sufficiently close to the care_state.csv data which was collected in 2023 and the beginning part of 2024. I plan to answer this question visually by first introducing how many states are in each bucket of median % ER departures, which ranges from 1-5%. I then will go on to compare population by state to this median ER departure % to determine if the state population leads to an effect on this departure % metric.

## Weekly Plan

Week 1 - Pick a data set and begin studying the data

Week 2 - Begin writing proposal

Week 3 - Finalize questions and analysis plan

Week 4 - Create visuals and work on formatting

Week 4-5 - Peer reviews and close open issues

Week 6 - Implement instructor feedback

Week 6-7 - Ensure project is reproducible, complete presentation, and complete write-up
