---
title: "Emergency Room Care Across the United States"
subtitle: "INFO 526 Data Analysis and Visualization"
author: 
  - name: "Michael Burris"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "Final project for INFO 526 Data Analysis and Visualization"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false

# Load packages
#install.packages("pacman")
pacman::p_load(tidyverse,
               scales,
               ggthemes
               )

# General settings
# Theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# Code output width
options(width = 65)

# Figure parameters
knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.68,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)

```

## Data Set

```{r}
#| label: load-dataset
#| message: false
# Load dataset
er_data <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-08/care_state.csv")
glimpse(er_data)

```

This data set comes from the Centers for Medicare and Medicaid Services (CMS) website. There are 1,232 rows and 8 columns of data. I chose this data set because I'm interested in working full-time in a Biotechnology or Healthcare-related industry, and this project presents a perfect opportunity for me to visualize and analyze health data.

## Questions

*The two questions you want to answer.*

1.  Which states/regions are the most and least effective time wise at treating psychiatric/mental health patients that are admitted to the emergency room?
2.  Does state population have an effect on what percent of patients leave the emergency room without being seen?

## Analysis plan

*A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).*

- To answer which states/regions are the most and least effective time wise at treating psychiatric/mental health patients admitted to the emergency room, I plan to create two visuals. I will need to use various variables from the care_state.csv file including condition, score, and measure_id. I will also need to utilize some external data that specifies the region for key states, states that have the shortest and longest median ER visit times. I will first visualize an overview of which states have the highest and lowest median visit times, then I will go a step further and visualize this by region as well.
- In order to address if state population has an effect on the percent of patients who leave the emergency room without being seen, I will need to use the condition, measure_id, and score variables. I will also need to pull in some external data to get population numbers for each state. I found an external source with population data estimates for each state for 2025, which I felt was sufficiently close to the care_state.csv data which was collected in 2023. I plan to answer this question visually by first introducing how many states are in each bucket of median % ER departures, which ranges from 1-5%. I then will go on to compare population by state to this median ER departure % to determine if the state population leads to an effect on this departure % metric.

## Visuals

```{r}
#| label: Answer to question 1

# Filter data to only include emergency room observations, patients that are non psyciatric/mental health patients, and only states not federal districts or territories.
patients <- er_data |>
  filter(condition == "Emergency Department", 
         measure_id == "OP_18c", state != "DC", state != "AS", state != "GU",
         state != "MP", state != "PR", state != "VI")

# Create data frame that only includes the 10 states with the longest ER visit times
longest_visits <- patients |>
  slice_max(score, n = 5)

# Create a data frame that only includes the 10 states with the shortest ER visit times
shortest_visits <- patients |>
  slice_min(score, n = 5)

# Join the two new data frames together
p_data = longest_visits |>
  full_join(shortest_visits)

# Create a new column that spells out the state names that meet the above criteria
p_data =p_data |>
  mutate(
    state_full = case_when(
      state %in% c("DE") ~ "Deleware", state %in% c("KS") ~ "Kansas",
      state %in% c("MA") ~ "Massachusets", state %in% c("MD") ~ "Maryland",
      state %in% c("MT") ~ "Montana", state %in% c("ND") ~ "North Dakota",
      state %in% c("NE") ~ "Nebraska", state %in% c("NY") ~ "New York",
      state %in% c("RI") ~ "Rhode Island", state %in% c("SD") ~ "South Dakota"
    )
  )

# Plot my data
ggplot(p_data, aes(x = score, y = factor(state_full, levels = rev(levels(factor(state_full)))))) + 
  geom_point(size = 4) + 
  labs(
    title = "Highest & Lowest Median ER Visit Times",
    subtitle = "Psychiatric/mental health patients",
    x = "Minutes",
    y = NULL,
    caption = "Source: Centers for Medicare & Medicaid Services | TidyTuesday"
  ) + 
  theme(
    plot.subtitle = element_text(color = "cornsilk4"),
    plot.caption = element_text(color = "cornsilk4")
  )
```

```{=html}
<details>
<summary>Sources</summary>

https://dplyr.tidyverse.org/reference/filter.html
https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States
https://dplyr.tidyverse.org/reference/top_n.html
https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/filter
https://dplyr.tidyverse.org/reference/pull.html#ref-examples
https://dplyr.tidyverse.org/reference/slice.html#ref-examples
https://ggplot2.tidyverse.org/reference/geom_bar.html
https://www.rdocumentation.org/packages/data.table/versions/1.18.4/topics/merge
https://d2l.arizona.edu/d2l/le/content/1755726/viewContent/20559504/View
https://r-graph-gallery.com/267-reorder-a-variable-in-ggplot2.html
https://www.statology.org/ggplot-order-y-axis-alphabetically/
https://medium.com/@MattOldach_65321/reversing-the-order-of-axis-in-a-ggplot2-scatterplot-b1d729b89118
https://ggplot2.tidyverse.org/reference/geom_point.html#ref-examples
https://en.wikipedia.org/wiki/List_of_regions_of_the_United_States
https://kpuka.ca/resources/quarto_intro.html#sec-html_folding

</details>
```

This visual highlights the 5 states that have the shortest and longest median ER visit times for psychiatric/mental health patients. For visual 2, take it a step further and visualize region along with it.

```{r}
#| label: Answer to question 1 (cont.)

# Modify my data to include region for more detailed analysis
p_data <- p_data |>
  mutate(
    region = case_when(
      state %in% c("DE", "MD") ~ "South",
      state %in% c("KS", "ND", "NE", "SD") ~ "Midwest",
      state %in% c("MT") ~ "West",
      TRUE ~ "Northeast"
    )
  )

# Plot my data
ggplot(p_data, aes(x = score, y = sort(state_full), fill = region)) + 
  geom_col() + 
  scale_fill_colorblind() + 
  labs(
    title = "Median ER Visit Times by State & Region",
    subtitle = "Psychiatric/mental health patients",
    x = "Minutes",
    y = NULL,
    caption = "Source: Centers for Medicare & Medicaid Services | TidyTuesday"
  ) + 
  theme(
    plot.subtitle = element_text(color = "cornsilk4"),
    plot.caption = element_text(color = "cornsilk4")
  )
```

```{=html}
<details>
<summary>Sources</summary>

https://ggplot2.tidyverse.org/reference/geom_bar.html#ref-examples
https://www.datacamp.com/tutorial/sorting-in-r
https://ggplot2.tidyverse.org/reference/geom_bar.html
https://forum.posit.co/t/changing-position-of-plot-captions/108973/2

</details>
```

This second visual clearly shows that the Midwest region offers some of the fastest ER visits while states in the South and Northeast offer the highest overall ER visits.

```{r}
#| label: Answer to question 2

# Filter data to only include % of patients who left the ER without being seen in each state
p_left <- er_data |>
  filter(condition == "Emergency Department",
         measure_id == "OP_22", state != "DC", state != "AS", state != "GU", 
         state != "MP", state != "PR", state != "VI")

# Load in new state population data
state_populations <- read_csv("data/state_populations.csv")

# Left join
combined_data <- left_join(p_left, state_populations)

# Set my theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# Plot the data
ggplot(combined_data, aes(x = score)) + 
         geom_histogram(bins = 5) + 
  scale_x_continuous(labels = label_percent(scale = 1)) + 
  labs(
    title = "Count of States and their Early ER Departures",
    x = NULL,
    y = "# of states",
    caption = "Source: Centers for Medicare & Medicaid Services | TidyTuesday"
  ) + 
  theme(
    plot.caption = element_text(color = "cornsilk4")
  )

```

```{=html}
<details>
<summary>Sources</summary>

https://ggplot2.tidyverse.org/reference/scale_continuous.html

</details>
```

This histogram shows the number of states that have 1%, 2%, 3%. 4%, and 5% median patients who left the ER prematurely.

```{r}
#| label: Answer to question 2 (cont.)

# Plot the data
ggplot(combined_data, aes(x = score, y = population, size = population, color = population)) + 
  #geom_point() + 
  geom_jitter() +
  scale_x_continuous(label = label_percent(scale = 1)) + 
  scale_y_log10(labels = scales::label_number()) +
  #scale_y_continuous(label = label_number()) + 
  labs(
    title = "Median % Early ER Departures by State",
    caption = "Source: Centers for Medicare & Medicaid Services | TidyTuesday",
    x = NULL,
    y = "State Population"
  ) + 
  guides(
    title = "State Population",
    size = "none",
    color = "none"
    ) + 
  theme(
    plot.caption = element_text(color = "cornsilk4")
  )
```

```{=html}
<details>
<summary>Sources</summary>

https://www.statsamerica.org/sip/rank_list.aspx?rank_label=pop1
https://www.geeksforgeeks.org/r-language/transform-ggplot2-plot-axis-to-log-scale-in-r/
https://ggplot2.tidyverse.org/articles/faq-axes.html
https://r-graph-gallery.com/79-levelplot-with-ggplot2.html
https://r-graph-gallery.com/320-the-basis-of-bubble-plot.html
https://d2l.arizona.edu/d2l/le/content/1755726/viewContent/20559504/View
https://ggplot2.tidyverse.org/reference/scale_continuous.html
https://www.r-bloggers.com/2021/08/beginning-a-ggplot2-series-logarithmize-your-scales/#google_vignette
https://ggplot2.tidyverse.org/reference/geom_jitter.html
https://ggplot2.tidyverse.org/reference/scale_size.html
https://ggplot2.tidyverse.org/reference/guide_legend.html
https://www.r-bloggers.com/2024/04/guide-to-remove-legends-in-ggplot2-in-r-programming/#google_vignette
https://www.r-bloggers.com/2024/04/guide-to-remove-legends-in-ggplot2-in-r-programming/
https://ggplot2.tidyverse.org/reference/geom_abline.html
https://www.datacamp.com/tutorial/r-median-function

</details>
```

It does not seem that population has an effect on the median percent of early ER departures. The distribution of states with smaller, medium, and larger sized populations are distributed fairly evenly across the median percent rates (1-5%).
