---
title: "Final Project"
subtitle: "INFO 526 Data Analysis and Visualization"
author: 
  - name: "Michael Burris"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "Final project for INFO 526 Data Analysis and Visualization"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

## Questions

The two questions you want to answer.

## Analysis plan

- A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
